Global study of well-being and internet adoption | Online analysis supplement

### 2021-11-22 03-model-estimate.Rmd
Refactor model fitting code

```{r data-prepare}
# Predictors

# Transform edge proportions to fit beta distribution
itu <- itu %>% 
  mutate(internet = if_else(internet == 0, .0000001, internet)) %>% 
  mutate(internet = if_else(internet == 1, .9999999, internet))

# Create lagged predictors
itu <- itu %>% 
  arrange(country, year) %>% 
  group_by(country) %>% 
  mutate(
    across(
      c(internet, fixed, mobile), 
      list(`1` = ~lag(., 1)), 
      .names = "{str_sub(.col, 1, 1)}{.fn}"
    )
  )

# Centering within and between countries
itu <- itu %>%
  group_by(country) %>% 
  mutate(
    i1_cb = mean(i1, na.rm = TRUE), # Country average
    i1_cw = i1 - i1_cb # Year-specific deviation from country average
  ) %>% 
  ungroup() %>% 
  # Grand-mean center the country averages
  mutate(i1_cb = i1_cb - mean(i1, na.rm = TRUE))

# Outcomes

# Merge predictors to outcomes
dat <- left_join(gbd, itu)

# Center and scale year, and contrast code factors
dat <- dat %>% 
  mutate(year = (year - 2010) / 10)
dat$sex <- factor(dat$sex)
contrasts(dat$sex) <- "contr.sum"
dat$age <- factor(dat$age)
contrasts(dat$age) <- "contr.sum"

# For brms multivariate data subsetting: variable to indicate unique ITU values
dat <- dat %>% 
  group_by(country, year) %>% 
  mutate(itu = 1:n() == 1) %>% 
  ungroup()
```

All the outcome variables are proportions (0-1) at this point. They need to be rescaled for HMC efficiency (self-harm models do not converge at this scale.) So remember to back-transform when interpreting results!

```{r}
# Save this so its easy to use later to back-transform
dat %>%
  group_by(cause) %>%
  summarise(median = median(val), min = min(val), max = max(val))

# Important: do this for both val AND se
dat <- dat %>% 
  group_by(cause) %>% 
  mutate(across(val:se, ~.x*100)) %>% 
  ungroup()

dat %>%
  group_by(cause) %>%
  summarise(median = median(val), min = min(val), max = max(val))
```

@@

### Model 1a

@@

Internet adoption is treated as Beta-distributed. Outcome rates, the GBD estimates, are treated as Gaussian with a known standard error, resulting in a "meta-regression" model of the outcomes.

$$
\begin{align*}
y_{i[c][r]} &\sim \text{Normal}(\mu^y_{i[c][r]}, \sigma^{2y}v_i) \\

@@

\left[\begin{array}{c}
\delta^y_{0[r]} \\ \delta^y_{1[r]} \\ \delta^x_{0[r]} \\ \delta^x_{1[r]}
\end{array}\right] 
\sim MVN(
\left[\begin{array}{c}
0 \\ 0 \\ 0 \\ 0
\end{array}\right],
\Sigma^r
)
\end{align*}
$$

@@

```{r}
# Define models of outcomes
bf_1a <- bf(
  val | resp_se(se, sigma = TRUE) ~ 
    year + 
    (year |c| country) +
    (year |r| region),
  family = gaussian()
)
```

We then compile the model once, such that it can be updated with subsets of the data (the different outcomes) without having to compile again.

```{r}
fit_1a <- brm(
  bf_itu + bf_1a + set_rescor(FALSE),
  data = filter(dat, cause == "Anxiety"),
  chains = 0,
  threads = hmc$threads,
  file = "output/brm-1a"
)
```

Then fit the model. Note we don't need this for now because the relevant estimates are provided by model 1c.

```{r eval = FALSE}
fits_1a <- dat %>% 
  group_by(cause) %>% 
  summarise(
    fit = list(
      update(
        fit_1a, 
        newdata = cur_data_all(),
        chains = hmc$chains, cores = hmc$cores, threads = hmc$threads,
        iter = hmc$iter, warmup = hmc$warmup,
        control = list(adapt_delta = hmc$d, max_treedepth = hmc$m),
        file = str_glue("output/brm-1a-{cur_group()}")
      )
    )
  )
```

### Model 1b

Model 1b is as 1a but we include age (10 to 14, 15 to 19, 20 to 24) and sex as covariates.

```{r}
bf_1b <- bf(
  val | resp_se(se, sigma = TRUE) ~ 
    year + sex + age + 
    (year + sex + age |c| country) +
    (year + sex + age |r| region),
  family = gaussian()
)
```

Then fit the model as above, first compiling and then updating for each outcome. Note not fitted for now--relevant estimates are in model 1c.

```{r eval = FALSE}
fit_1b <- brm(
  bf_itu + bf_1b + set_rescor(FALSE),
  data = filter(dat, cause == "Anxiety"),
  chains = 0,
  threads = hmc$threads,
  file = "output/brm-1b"
)

fits_1b <- dat %>% 
  group_by(cause) %>% 
  summarise(
    fit = list(
      update(
        fit_1b, 
        newdata = cur_data_all(),
        chains = hmc$chains, cores = hmc$cores, threads = hmc$threads,
        iter = hmc$iter, warmup = hmc$warmup,
        control = list(adapt_delta = hmc$d, max_treedepth = hmc$m),
        file = str_glue("output/brm-1b-{cur_group()}")
      )
    )
  )
```

### Model 1c

Model 1c further expands 1b to include the two-way interactive terms between age and year, and sex and year. Thus we get age and sex-specific trajectories over time. We reuse the prior distributions from 1b.

This takes about 3.5 hours per 1000 iterations (per model).

```{r}
bf_1c <- bf(
  val | resp_se(se, sigma = TRUE) ~ 
    year * sex + year * age + 
    (year * sex + year * age |c| country) +
    (year * sex + year * age |r| region),
  family = gaussian()
)

# Compile model separately for each outcome to give independent default priors
fits_1c <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit = list(
      brm(
        formula = bf_itu + bf_1c + set_rescor(FALSE), 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-1c-{cur_group()}")
      )
    )
  )

bf_1c_no_se <- bf(
  val ~ 
    year * sex + year * age + 
    (year * sex + year * age |c| country) +
    (year * sex + year * age |r| region),
  family = gaussian()
)

# Compile model separately for each outcome to give independent default priors
fits_1c <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit_no_se = list(
      brm(
        formula = bf_itu + bf_1c_no_se + set_rescor(FALSE), 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-1c-no-se-{cur_group()}")
      )
    )
  )
```

@@

```{r}
f1d <- fits_1c %>% 
  mutate(
    out = 
      map(
        fit, 
        ~as_draws_df(., variable = "^b_|^sd_|^cor_", regex = TRUE) %>% 
          summarise_draws(samples = length, default_convergence_measures())
      )
  ) %>% 
  select(-fit) %>% 
  unnest(out)
f1d %>%  
  kbl(
    caption = "HMC Diagnostics, Model 1c",
    digits = c(0,0,0,2,0,0)
  ) %>% 
  kable_paper(html_font = Font) %>% 
  column_spec(
    4, 
    color = "white",
    background = spec_color(
      f1d$rhat, 
      direction = 1, 
      option = "C", 
      end = .7
    )
  )
```

```{r}
tmp <- fits_1c %>% 
  mutate(
    out = map(
      fit, 
      ~pp_check(
        .x, 
        ndraws = 10, 
        type = "dens_overlay_grouped",
        group = "region",
        resp = "val", 
        newdata = .x$data
      ) + facet_wrap("group", nrow = 1, scales = "free")
    )
  )
wrap_plots(tmp$out, nrow = 3) &
  scale_x_continuous(
    breaks = pretty_breaks(2),
    labels = function(x) percent(x/100)
  )
```

@@

@@

```{r}
bf_2b <- bf(
  val | resp_se(se, sigma = TRUE) ~ 
    year + sex + age + i1_cw + i1_cb +
    (year + sex + age + i1_cw | country) +
    (year + sex + age + i1_cw | region),
  family = gaussian(), 
  unused = ~ cause # Save this variable in the data frame
)
fit_2b <- brm(
  bf_2b,
  data = filter(dat, cause == "Anxiety"),
  chains = 0,
  threads = hmc$threads,
  file = "output/brm-2b"
)
fits_2b <- dat %>% 
  group_by(cause) %>% 
  summarise(
    fit = list(
      update(
        fit_2b, 
        newdata = cur_data_all(),
        chains = hmc$chains, cores = hmc$cores, threads = hmc$threads,
        iter = hmc$iter, warmup = hmc$warmup, refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-2b-{cur_group()}")
      )
    )
  )

bf_2c <- bf(
  val | se(se, sigma = TRUE) ~ 
    year * sex + year * age + i1_cw * sex + i1_cw * age + i1_cb + 
    (year * sex + year * age + i1_cw * sex + i1_cw * age | country) +
    (year * sex + year * age + i1_cw * sex + i1_cw * age | region),
  family = gaussian()
)

bf_2c_no_se <- bf(
  val ~ 
    year * sex + year * age + i1_cw * sex + i1_cw * age + i1_cb + 
    (year * sex + year * age + i1_cw * sex + i1_cw * age | country) +
    (year * sex + year * age + i1_cw * sex + i1_cw * age | region),
  family = gaussian()
)

# Compile model separately for each outcome to give independent default priors
fits_2c <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit = list(
      brm(
        formula = bf_2c, 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-2c-{cur_group()}")
      )
    )
  )

fits_2c <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit_no_se = list(
      brm(
        formula = bf_2c_no_se, 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-2c-no-se-{cur_group()}")
      )
    )
  )
```

@@

```{r}
f2d <- fits_2c %>% 
  mutate(
    out = 
      map(
        fit, 
        ~as_draws_df(., variable = "^b_|^sd_", regex = TRUE) %>% 
          summarise_draws(samples = length, default_convergence_measures())
      )
  ) %>% 
  select(-fit) %>% 
  unnest(out)
f2d %>%  
  kbl(
    caption = "HMC Diagnostics, Model 2c",
    digits = c(0,0,0,2,0,0)
  ) %>% 
  kable_paper(html_font = Font, full_width = FALSE) %>% 
  column_spec(
    4, 
    color = "white",
    background = spec_color(
      f2d$rhat, 
      direction = 1, 
      option = "C", 
      end = .7
    )
  ) %>% 
  scroll_box(height = "600px", width = "800px")
```

```{r}
tmp <- fits_2c %>% 
  mutate(
    out = map(
      fit, 
      ~pp_check(
        .x, 
        ndraws = 10, 
        type = "dens_overlay_grouped",
        group = "region",
        resp = "val", 
        newdata = .x$data
      ) + 
        facet_wrap("group", nrow = 1, scales = "free_y")
    )
  )
wrap_plots(tmp$out, nrow = 3) &
  scale_x_continuous(
    breaks = pretty_breaks(2),
    labels = function(x) percent(x/100, 1)
  )
```

@@

```{r}
tmp <- fits_2c %>% 
  mutate(
    out = map(
      fit, 
      ~as_draws_df(.x, variable = "^b_", regex = TRUE) %>% 
        summarise_draws(
          mean, quantile, rhat, 
          .args = list(probs = c(.025, .975))
        )
    )
  ) %>% 
  select(-fit) %>% 
  unnest(out) %>% 
  mutate(across(where(is.numeric), ~round(., 2))) %>% 
  mutate(
    result = str_glue(
      "b = {mean} ({`2.5%`}, {`97.5%`})"
    )
  ) %>% 
  rowwise() %>% 
  select(-c(mean, `2.5%`, `97.5%`, rhat)) %>% 
  pivot_wider(names_from = cause, values_from = result)
tmp %>%   
  kbl() %>% 
  kable_paper(html_font = Font) 
```

@@

```{r}
tmp <- fits_2c %>% 
  mutate(
    out = map(
      fit, 
      ~emtrends(.x, var = "i1_cw", by = c("sex", "age"))
    )
  ) %>% 
  select(-fit)
tmp %>% 
  mutate(out = map(out, ~as.data.frame(out))) %>% 
  unnest(out) %>% 
  select(cause, sex, age, i1_cw.trend, lower.HPD, upper.HPD) %>% 
  kbl(digits = 3) %>% 
  kable_minimal(full_width = FALSE)
```

>>>>

```{r data-prepare}
# Merge predictors to outcomes
dat <- left_join(gbd, itu)

# Center and scale year, and contrast code factors
dat <- dat %>% 
  mutate(year = (year - 2010) / 10)
dat$sex <- factor(dat$sex)
contrasts(dat$sex) <- "contr.sum"
dat$age <- factor(dat$age)
contrasts(dat$age) <- "contr.sum"
```

All the outcome variables are proportions (0-1) at this point. We rescale them to percentages (0-100).

```{r}
# Important: do this for both val AND se
dat <- dat %>% 
  mutate(across(val:se, ~.x*100))
```

@@

### Model 1

@@

Internet adoption is treated as Beta-distributed. Outcome rates are treated as gaussian.

Prepare data for this model

```{r}
dat_model1 <- dat %>% 
  # Transform edge proportions to fit beta distribution
  mutate(internet = if_else(internet == 0, .0000001, internet)) %>% 
  mutate(internet = if_else(internet == 1, .9999999, internet)) %>% 
  # Indicate unique ITU values for MV model...
  arrange(region, country, year, cause, sex, age) %>% 
  # ...Use only the first internet value within a year-country
  group_by(region, country, year) %>% 
  mutate(itu = 1:n() == 1) %>% 
  ungroup()
```

$$
\begin{align*}
y_{i[c][r]} &\sim \text{Normal}(\mu^y_{i[c][r]}, \sigma^{2y}) \\

@@

\left[\begin{array}{c}
\delta^y_{0[r]} \\ \delta^y_{1[r]} \\ \delta^x_{0[r]} \\ \delta^x_{1[r]}
\end{array}\right] 
\sim MVN(
\left[\begin{array}{c}
0 \\ 0 \\ 0 \\ 0
\end{array}\right],
\Sigma^r
)
\end{align*}
\mathbf{\beta}
$$

@@

```{r}
# Define models of outcomes
bf_1 <- bf(
  val ~ 
    year * sex * age + 
    (year * sex * age |c| country) +
    (year * sex * age |r| region),
  family = gaussian()
)

bf_1_se <- bf(
  val | resp_se(se, sigma = TRUE) ~ 
    year * sex * age + 
    (year * sex * age |c| country) +
    (year * sex * age |r| region),
  family = gaussian()
)
```

```{r}
fits_1 <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit = list(
      brm(
        formula = bf_itu + bf_1 + set_rescor(FALSE), 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-1-{cur_group()}")
      )
    )
  )

fits_1 <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit_se = list(
      brm(
        formula = bf_itu + bf_1_se + set_rescor(FALSE), 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-1-se-{cur_group()}")
      )
    )
  )
```

@@

```{r}
f1d <- fits_1 %>% 
  mutate(
    out = 
      map(
        fit, 
        ~as_draws_df(., variable = "^b_|^sd_|^cor_", regex = TRUE) %>% 
          summarise_draws(samples = length, default_convergence_measures())
      )
  ) %>% 
  select(-fit) %>% 
  unnest(out)
f1d %>%  
  kbl(
    caption = "HMC Diagnostics, Model 1c",
    digits = c(0,0,0,2,0,0)
  ) %>% 
  kable_paper(html_font = Font) %>% 
  column_spec(
    4, 
    color = "white",
    background = spec_color(
      f1d$rhat, 
      direction = 1, 
      option = "C", 
      end = .7
    )
  )
```

```{r}
tmp <- fits_1 %>% 
  mutate(
    out = map(
      fit, 
      ~pp_check(
        .x, 
        ndraws = 10, 
        type = "dens_overlay_grouped",
        group = "region",
        resp = "val", 
        newdata = .x$data
      ) + facet_wrap("group", nrow = 1, scales = "free")
    )
  )
wrap_plots(tmp$out, nrow = 3) &
  scale_x_continuous(
    breaks = pretty_breaks(2),
    labels = function(x) percent(x/100)
  )
```

@@

Prepare data for this model

```{r}
# Create lagged predictors
dat <- dat %>% 
  arrange(region, country, cause, sex, age, year) %>% 
  group_by(region, country, cause, sex, age) %>% 
  mutate(
    across(
      c(internet, fixed, mobile), 
      list(`1` = ~lag(., 1)), 
      .names = "{str_sub(.col, 1, 1)}{.fn}"
    )
  )

# Centering within and between countries
dat <- dat %>%
  group_by(country) %>% 
  mutate(
    i1_cb = mean(i1, na.rm = TRUE), # Country average
    i1_cw = i1 - i1_cb # Year-specific deviation from country average
  ) %>% 
  ungroup() %>% 
  # Grand-mean center the country averages
  mutate(i1_cb = i1_cb - mean(i1, na.rm = TRUE))
```

@@

```{r}
bf_2 <- bf(
  val ~ 
    year * sex + year * age + i1_cw * sex + i1_cw * age + i1_cb + 
    (year * sex + year * age + i1_cw * sex + i1_cw * age | country) +
    (year * sex + year * age + i1_cw * sex + i1_cw * age | region),
  family = gaussian()
)

bf_2_se <- bf(
  val | se(se, sigma = TRUE) ~ 
    year * sex * age + i1_cw * sex * age + i1_cb + 
    (year * sex * age + i1_cw * sex * age | country) +
    (year * sex * age + i1_cw * sex * age | region),
  family = gaussian()
)

# Compile model separately for each outcome to give independent default priors
fits_2 <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit = list(
      brm(
        formula = bf_2, 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-2-{cur_group()}")
      )
    )
  )

fits_2 <- dat %>%
  group_by(cause) %>% 
  summarise(
    fit_se = list(
      brm(
        formula = bf_2_se, 
        data = cur_data_all(),
        chains = hmc$chains, 
        cores = hmc$cores, 
        threads = hmc$threads,
        iter = hmc$iter, 
        warmup = hmc$warmup, 
        refresh = hmc$refresh,
        control = list(
          adapt_delta = hmc$adapt_delta, 
          max_treedepth = hmc$max_treedepth
        ),
        file = str_glue("output/brm-2-se-{cur_group()}")
      )
    )
  )
```

@@

```{r}
f2d <- fits_2 %>% 
  mutate(
    out = 
      map(
        fit, 
        ~as_draws_df(., variable = "^b_|^sd_", regex = TRUE) %>% 
          summarise_draws(samples = length, default_convergence_measures())
      )
  ) %>% 
  select(-fit) %>% 
  unnest(out)
f2d %>%  
  kbl(
    caption = "HMC Diagnostics, Model 2c",
    digits = c(0,0,0,2,0,0)
  ) %>% 
  kable_paper(html_font = Font, full_width = FALSE) %>% 
  column_spec(
    4, 
    color = "white",
    background = spec_color(
      f2d$rhat, 
      direction = 1, 
      option = "C", 
      end = .7
    )
  ) %>% 
  scroll_box(height = "600px", width = "800px")
```

```{r}
tmp <- fits_2 %>% 
  mutate(
    out = map(
      fit, 
      ~pp_check(
        .x, 
        ndraws = 10, 
        type = "dens_overlay_grouped",
        group = "region",
        resp = "val", 
        newdata = .x$data
      ) + 
        facet_wrap("group", nrow = 1, scales = "free_y")
    )
  )
wrap_plots(tmp$out, nrow = 3) &
  scale_x_continuous(
    breaks = pretty_breaks(2),
    labels = function(x) percent(x/100, 1)
  )
```

@@

```{r}
tmp <- fits_2 %>% 
  mutate(
    out = map(
      fit, 
      ~as_draws_df(.x, variable = "^b_", regex = TRUE) %>% 
        summarise_draws(
          mean, quantile, rhat, 
          .args = list(probs = c(.025, .975))
        )
    )
  ) %>% 
  select(-fit) %>% 
  unnest(out) %>% 
  mutate(across(where(is.numeric), ~round(., 2))) %>% 
  mutate(
    result = str_glue(
      "b = {mean} ({`2.5%`}, {`97.5%`})"
    )
  ) %>% 
  rowwise() %>% 
  select(-c(mean, `2.5%`, `97.5%`, rhat)) %>% 
  pivot_wider(names_from = cause, values_from = result)
tmp %>%   
  kbl() %>% 
  kable_paper(html_font = Font) 
```

@@

```{r}
tmp <- fits_2 %>% 
  mutate(
    out = map(
      fit, 
      ~emtrends(.x, var = "i1_cw", by = c("sex", "age"))
    )
  ) %>% 
  select(-fit)
tmp %>% 
  mutate(out = map(out, ~as.data.frame(out))) %>% 
  unnest(out) %>% 
  select(cause, sex, age, i1_cw.trend, lower.HPD, upper.HPD) %>% 
  kbl(digits = 3) %>% 
  kable_minimal(full_width = FALSE)
```
